---
title: "Exercises Day 1 Part 2"
author: "Bhavya Kaushik"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the data
Question: Load the gapminder dataset.

I am using the read_csv function from readr package to read my data.

```{r load the data}
library(readr)
gapminder <- read_csv(here::here("data", "gapminder5.csv"))
head(gapminder, n = 5)
```

## If Statement 

Question: Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. 

Hint: use the `any` function.

```{r if statement}
year<-2002
if(any(gapminder$year == year)){
   print(paste("Record(s) for the year",year,"found."))
} else {
  print(paste("No records for year",year))
}
```

Now do the same for 2012.

```{r if statement 2}
year<-2012
if(any(gapminder$year == year)){
   print(paste("Record(s) for the year",year,"found."))
} else {
  print(paste("No records for year",year))
}
```

## Loop and If Statements

Question: Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

```{r Loop and If Statements}
overall_mean = mean(gapminder$pop)
for (i in unique(gapminder$country)) {
   country_mean = mean(gapminder$pop[gapminder$country==i])
   
   if (country_mean < overall_mean) {
     mean_le = mean(gapminder$lifeExp[gapminder$country==i])
     print(paste("Mean Life Expectancy in", i, "is", mean_le))
   } 
} # end for loop
```

Question: Write a script that loops through the `gapminder` data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{r Loop and If Statements 2}
lower_threshold <- 50
upper_threshold <- 70
 
for (i in unique(gapminder$continent)){
    tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
    
    if (tmp < lower_threshold){
        print(paste("Average Life Expectancy in", i, "is less than", lower_threshold))
    }
    else if (tmp > lower_threshold & tmp < upper_threshold){
        print(paste("Average Life Expectancy in", i, "is between", lower_threshold, "and", upper_threshold))
    }
    else {
        print(paste("Average Life Expectancy in", i, "is greater than", upper_threshold))
    }
   
}
```

## Writing Functions

Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset.  Hint: Use `mode()` or `class()` to get the class of the data in each column. Remember that `names()` or `colnames()` returns the name of the columns in a dataset.

```{r Writing Functions}
data_frame_info <- function(df) {
  cols <- names(df)
  for (i in cols) {
    print(paste0(i, ": ", mode(df[, i])))
  }
}
data_frame_info(gapminder)
```

Create a function that given a vector will print the mean and the standard deviation of a **vector**, it will optionally also print the median. Hint: include an argument that takes a boolean (`TRUE`/`FALSE`) operator and then include an `if` statement.

```{r Writing Functions 2}
vector_info <- function(x, include_median=FALSE) {
  print(paste("Mean:", mean(x)))
  print(paste("Standard Deviation:", sd(x)))
  if (include_median) {
    print(paste("Median:", median(x)))
  }
}

le <- gapminder$lifeExp
vector_info(le, include_median = F)
vector_info(le, include_median = T)
```

## Analyzing the relationship between GDP per capita and life expectancy

Use what you've learned so far to answer the following questions using the `gapminder` dataset. Be sure to include some visualizations!

#### Question 1. What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)

```{r Analyzing}
# Write a function to plot the linear natural log-log model relating GDP per capita to life expectancy 
viz_lm <-
    function(df, dv, iv, year) {
        dat <- df[df[["year"]] == year, ]
        y <- log(dat[[dv]])
        x <- log(dat[[iv]])
        fit <- lm(y ~ x)
        plot(y ~ x, main = year,
             xlab = "log(lifeExp)", ylab = "log(gdpPercap)")
        lines(x, predict(fit), col = 'blue')
        print(paste("intercept obtained after fitting linear model =", coef(fit)[1]))
        print(paste("slope obtained after fitting linear model =", coef(fit)[2]))
    }
```

```{r Analyzing 2}
# Make a graph:
viz_lm(gapminder, "gdpPercap", "lifeExp", 1977)
```

```{r Analyzing 3}
# loop it
years = unique(gapminder$year)
for (i in years) {
    viz_lm(gapminder, "gdpPercap", "lifeExp", i)
}
```
The relationship is positive. GDP per capita is increasing with an increase in life expectancy.
The magnitude of the slope has followed an increasung trend from 1952 to 1987. After 1987 it decreases slightly and kind of plateaus.


#### Question 2. Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.

```{r Analyzing 4}

```